---
title: "differential_outcomes"
author: "Brian M. Schilder"
date: "2023-03-23"
output: html_document
---

```{r setup, fig.height=10, fig.width=8}
library(data.table)
library(ggplot2)
library(ggridges)
```


> Can we say, that this phenotype, when it acts via this celltype, has a different clinical course?

# Import data

```{r}
d <- MultiEWCE::load_example_results("Descartes_All_Results_extras.symptoms.full_join.rds")
annot <- HPOExplorer::make_phenos_dataframe(
  add_hoverboxes = FALSE,
  add_disease_data = TRUE)
```


Find phenotypes associated w/ 20+ diseases & have metadata.

Then test whether the EWCE enrichment *p-values* (conditional on **celltype**) significantly affect whether the **phenotype** affects clinical outcomes (e.g. age of death).

The formula for the linear model is as follows:
> `outcome ~ EWC_pvalue * celltype`

# Run tests

```{r Run tests}  
save_path <- here::here("reports","differential_outcomes.csv.gz")
if(file.exists(save_path)){
  res_dt <- data.table::fread(save_path)
} else {
    vars <- c("AgeOfDeath_score",
              "Onset_score",
              "Severity_score","tier_merge",
              "pheno_freq_min","pheno_freq_max","pheno_freq_mean"
            # "symptoms.pval"
            )
  set.seed(2023)
  workers <- 10
  # future::plan("multicore",workers=workers) 
  options(future.globals.maxSize = 8000*1024^2)
  #### Iterate over each variable ####
  res_dt <- lapply(stats::setNames(vars,vars),
                   function(v){
    a1 <- annot[!is.na(get(v))]
    a1[,n_diseases:=length(unique(DatabaseID)), by="HPO_ID"]
    a1 <- a1[n_diseases>10]
    hpo_ids <- unique(a1$HPO_ID) |> head(n = 100)
    message("Testing: ",v," in ",length(hpo_ids)," HPO_IDs")
    #### Iterate over each HPO ID ####
    parallel::mclapply(hpo_ids,
                       mc.cores = workers,
                       function(hpo_id){
    # furrr::future_map(.x = hpo_ids, 
    #                   .progress = TRUE,
    #                   .f = function(hpo_id){ 
      HPOExplorer:::message_parallel("+ ",hpo_id)
      d1 <- d[HPO_ID==hpo_id][,DatabaseID:=LinkID]    
      #### Merge data with clinical annotations ####
      da2 <- data.table::merge.data.table(
        a1[,c("HPO_ID","DatabaseID",v),with=FALSE] |>
          unique(),
        d1, 
        allow.cartesian = TRUE,
        by = c("HPO_ID","DatabaseID")
      )[!is.na(get(v))]
      # da2 <- da2[q<0.05]
      ## Check that there's enough variation
      ## in the clinical variable and celltypes
      if(length(unique(da2[[v]]))<2 | 
         length(unique(da2[["CellType"]]))<2 ) {
        return(NULL)
      }     
      res <- lm(data = da2, 
                # formula = paste(v,"~ CellType")
                formula = paste(v,"~ symptoms.pval*CellType")
                )  |>
        broom::tidy() |>
        data.table::data.table() 
      res$variable <- v
      res[,HPO_ID:=hpo_id]
      res[,test_id:=paste(variable,HPO_ID,sep=".")]
      #### Add model input data metrics ####
      res[,mdata_rows:=nrow(da2)]
      res[,mdata_celltypes:=length(unique(da2$CellType))]
      res[,mdata_diseases:=length(unique(da2$DatabaseID))]
      res[,mdata_mean_value:=mean(da2[[v]])]
      res[,mdata_mean_value_modifier_celltype:=mean(da2[symptoms.pval<0.05][[v]])]
      res[,mdata_mean_value_nonmodifier_celltype:=mean(da2[symptoms.pval>=0.05][[v]])]
      res[,model_valence:=ifelse(estimate[term=="(Intercept)"]>=0,"+","-")]
      return(res)
    }) |> data.table::rbindlist(fill = TRUE)
  }) |> data.table::rbindlist(fill = TRUE)  
  data.table::fwrite(res_dt,save_path)
} 
```

# Plot 

## Intercepts

Plotting the q-value of the intercept of each test basically tells us the number of phenotypes in which celltype was a strong modifier of each clinical course.

```{r}
plot_variables <- function(res_dt, 
                           type="barplot",
                           scales="free"){
  dat <- res_dt[term=="(Intercept)"][,q.value:=stats::p.adjust(p = p.value,method = "fdr")]
  if(type=="violin"){ 
    ggplot(dat, aes(x=-log10(q.value), y=variable, fill=variable)) +
      geom_violin(show.legend = FALSE, color=alpha("black",.3), na.rm = TRUE) +
      geom_point(size=1, alpha=.1, show.legend = FALSE, na.rm = TRUE) +
      geom_vline(xintercept = -log10(0.05), linetype="dashed", alpha=0.5,color="blue") +
      annotate(x=-log10(0.05), y=-1.1,
               label=expression("| q-value<0.05" %->% ""),
               geom = "text", hjust=0,
               size=3,color=alpha("black",.5)) +
      coord_cartesian(ylim = c(0, 8), clip = "off") +
      theme_bw()
  } else {
    ggplot(dat, aes(x=-log10(q.value), 
                          fill=variable)) +
      geom_histogram(bins = 75, show.legend = FALSE) +
      geom_vline(xintercept = -log10(0.05), linetype="dashed", alpha=0.5, color="blue") +
      facet_wrap(facets = variable~., scales = scales) +
      scale_fill_viridis_d(option = "mako", begin = .25, end = .8) +
      theme_bw() +
      theme(strip.background = element_rect(fill="white"))
  }
}
```

### Violin plot

Summarised as a violin plot:
```{r}
vp <- plot_variables(res_dt = res_dt,
                     type = "violin")
vp
```

Let's cut off super significant values so that we can see the distributions better:
```{r}
vp + xlim(0,110)
```


### Histograms

We can see the distributions a bit better with histograms.

The vast majority of tests came back as significant. Meaning celltypes very have an impact on all clinical course variables.

```{r, fig.width=8}
plot_variables(res_dt = res_dt,
               type = "histogram")
```


## Celltypes

> What proportion of celltypes are each phenotype enriched for?

Finally, let's plot the p-values of each celltype within the model.
This shows that celltypes tend not to be indiscriminately associated with most phenotype.

In other words, celltypes are relatively specific to certain symptoms/phenotypes.


```{r}
plot_celltypes <- function(res_dt,
                           v=NULL,
                           x_var="-log10(p.value)",
                           scales="free_y"){
  
  dat <- res_dt[grepl("^CellType",term)][,term:=gsub("^CellType","",term)]
  if(!is.null(v)){
      dat <- dat[variable %in% v]
  } 
  gp <- ggplot(dat, 
              aes_string(x=x_var, 
                        fill="term",
                        color="term")) +
    geom_density(adjust=.9, position = "stack", color=alpha("cyan",.75), linewidth=.01, na.rm = TRUE) + 
    scale_fill_manual(values = pals::ocean.phase(n = length(unique(dat$term)))) + 
    theme_bw() + 
    theme(legend.key.size = unit(1, 'lines'), #change legend key size
      legend.key.height = unit(.5, 'lines'), #change legend key height
      legend.key.width = unit(.5, 'cm'), #change legend key width
      legend.title = element_text(size=6), #change legend title font size
      legend.text = element_text(size=5), 
      legend.background = element_blank(),
      strip.background = element_rect(fill="white"), 
      legend.direction = "horizontal",
     legend.position="bottom"
     ) +
    guides(fill=guide_legend(ncol=5)) + 
    facet_wrap(facets = variable~.,
               scales = scales) 

  if(x_var=="-log10(p.value)"){
      gp <- gp + 
        geom_vline(xintercept = -log10(0.05), linetype="dashed", alpha=0.5, color="blue")
  }
  return(gp)
}
```


```{r, fig.height=7, fig.width=10, warning=FALSE}
plot_celltypes(res_dt = res_dt)
```


> What is the magnitude of effect of celltype identity on each clinical course?


```{r}
plot_celltypes(res_dt = res_dt, 
               x_var = "log10(abs(estimate))", 
               scales="free")
```


## Distributions

Now let's highlight some specific examples where celltype can cause a differential clinical course.
 
```{r} 
sig_ids <- res_dt[term=="(Intercept)"][,q.value:=stats::p.adjust(p = p.value,method = "fdr")][q.value<0.05]$test_id

sig_dt <- res_dt[test_id %in% sig_ids][grepl("^CellType",term)][,CellType:=gsub("^CellType","",term)] 

message(formatC(length(unique(sig_dt$HPO_ID)),big.mark = ","),
        " phenotypes have a differential clinical course dependent on the celltype:") 
```


```{r}
# sig_dt_melt <- data.table::melt.data.table(
#   sig_dt,
#   measure.vars = c("mdata_mean_value",
#                    "mdata_mean_value_modifier_celltype",
#                    "mdata_mean_value_nonmodifier_celltype"), 
#   variable.name = "mdata_mean_variable", 
#   value.name = "mdata_mean_value")
# ggplot(sig_dt_melt, aes(x=mdata_mean_value,
#                         fill=mdata_mean_variable))+
#   geom_density() +
#   facet_wrap(facets = .~variable, scales = "free")
# 
# gp2 <- ggplot(sig_dt, aes(x=mdata_mean_value_nonmodifier_celltype))+
#   geom_density() +
#   facet_wrap(facets = .~variable, scales = "free")
  
```


## Compute directional differences

```{r}
id.vars <- c("HPO_ID","Phenotype","DatabaseID")
 
dannot <- data.table::merge.data.table(
  d[HPO_ID %in% unique(sig_dt$HPO_ID), DatabaseID:=LinkID],
  annot[,unique(c(id.vars,unique(sig_dt$variable)) ),with=FALSE],
  by = id.vars) 
d_melt <- 
  data.table::melt.data.table(
    dannot, 
    measure.vars = unique(sig_dt$variable))[!is.na(value)]


dat <- data.table::merge.data.table(sig_dt, 
                                    d_melt[,-c("intersection")],
                                    all.x = TRUE,
                                    # all.y = TRUE,
    by=c("HPO_ID","variable","CellType")) |> unique()
dat[is.na(p.value)]$p.value <- 1
dat[,modifying_celltype:=factor(p.value<.05, 
                                levels = c(TRUE, FALSE),
                                ordered = TRUE)]  
dat[,variable:=factor(variable, 
                      levels = unique(sig_dt$variable), 
                      ordered = TRUE)]
dat[,value_mean:=mean(value),by="variable"]
dat[,directional_diff:= ifelse(estimate>0,
                               mdata_mean_value + value,
                               mdata_mean_value - value)]
#### Make sure directional_diff predictions don't go outside the scale ####
# dat[,directional_diff:=ifelse(directional_diff<min(value),min(value),directional_diff), by="variable"]
# dat[,directional_diff:=ifelse(directional_diff>max(value),max(value),directional_diff), by="variable"]
dat[,directional_diff:=scales::rescale(directional_diff,c(min(value),max(value))),
    by="variable"]
# dat[,fdr:=stats::p.adjust(p.value,method = "fdr")]
```



### Top results

```{r}
top_dt <- sig_dt |>  
  subset(p.value<0.05) |>
  dplyr::group_by(variable) |>
  dplyr::slice_min(order_by = p.value,
                   n = 1) |>
  data.table::data.table()

MultiEWCE::create_dt(top_dt)
```

```{r}

top_dat <- dat[test_id %in% head(top_dt$test_id,4)]



plts <- lapply(stats::setNames(unique(top_dat_i$HPO_ID),
                       unique(top_dat_i$HPO_ID)), 
       function(id){
  top_dat_i <- top_dat[symptoms.pval<0.05 &
                         p.value<0.05 & 
                         HPO_ID == id]  
  top_dat_i[,label:=paste(sort(unique(DatabaseID)),
                                    collapse = "\n"),
            by=c("test_id","CellType")]
  top_dat_i[,group_mean:=mean(value),
            by=c("test_id","CellType")]
  top_dat_i <- top_dat_i |> 
    dplyr::arrange(dplyr::desc(group_mean)) |>
    dplyr::mutate(CellType=factor(CellType, unique(CellType),
                                  ordered = TRUE))|>
    data.table::data.table()
  top_dat_i[,n_diseases:=length(unique(DatabaseID)),
            by=c("test_id","CellType")]

  ggplot(top_dat_i, 
         aes(x=CellType,
             y=value,
             fill=n_diseases,
             label=label)) +
    geom_boxplot(show.legend = TRUE) +
    geom_point(show.legend = FALSE) +
    geom_line() + 
    scale_fill_viridis_c(option = "plasma") + 
    labs(title=top_dat_i$Phenotype[1],
         x=NULL,
         y=top_dat_i$variable[1]) + 
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust=1))
})
pw <- patchwork::wrap_plots(plts, ncol = 3, guides = "collect") 
pw
  
```




```{r}
ggplot(dat, aes(x = modifying_celltype,
                y= directional_diff , 
                fill = modifying_celltype)) +
  geom_violin(na.rm = TRUE) +
  geom_boxplot(fill="transparent", color="turquoise", 
               alpha=.2,
               outlier.alpha = .25, na.rm = TRUE) +
  facet_wrap(facets = .~variable, scales = "free") +
  scale_fill_viridis_d(option = "mako", begin = .25, direction = -1) +
  labs(x="Directional difference", y=NULL) +
  theme_bw() +
  theme(axis.text.x = element_blank(), 
        strip.background = element_blank())  
```



Let's plot the distribution each metadata attribute overall.

```{r} 
  ggplot(dat, 
         aes(x=value, 
             fill=variable,
             y = -0.5)) +
    geom_boxplot(position = position_dodge2(preserve = "single"), 
                 show.legend = FALSE, na.rm = TRUE) +
    geom_density(aes(x = value, fill = variable),
                 show.legend = TRUE, na.rm = TRUE, inherit.aes = FALSE,
                 alpha=.75, adjust=.5) + 
    facet_wrap(facets = variable~.,
               scales="free_x", 
               ncol = 3) +
    stat_boxplot(geom = "vline", aes(xintercept = ..xlower..),
                 width=position_dodge2(width = 1)) +
    stat_boxplot(geom = "vline", aes(xintercept = ..xmiddle..),
                 width=position_dodge2(width = 1)) +
    stat_boxplot(geom = "vline", aes(xintercept = ..xupper..),
                 width=position_dodge2(width = 1)) + 
    labs(y="Density") +
    scale_y_discrete(drop=FALSE) +
    scale_fill_viridis_d(option = "mako", alpha=.7,
                         begin = .25, end=.75,
                         drop=FALSE) + 
    theme_bw() + 
    theme(strip.background = element_blank())

```



Now let's try to project these difference back onto the full data distribution so we can see how different that looks on the scale.


We will want to differentiate between +/- values to get a sense of how much higher or how much lower celltypes can make a given clinical course.

If the estimate is negative, subtract the difference between the population mean and the values from the population mean:

When the estimates are positive 
> lower = mean + abs(mean-values)

When the estimates are negative
> lower = mean - abs(mean-values)

```{r, message=FALSE, warning=FALSE, fig.height=10, fig.width=13}   
plot_distributions <- function(dat,
                               d_melt,
                               value_var="directional_diff",
                               xlabel=value_var){
    
  lvls <- c("1. All",
            "2. Significant models\n(q-value<0.05)",
            "3. Non-modifying cell types\n(p-value>0.05)",
            "4. Modifying cell types\n(p-value<0.05)",
            "5. Modifying cell types\n(p-value<0.005)",
            "6. Modifying cell types\n(p-value<0.0001)",
            "7. Modifying cell types\n(p-value<0.0001 & estimate>0)") 
  lvls <- factor(lvls, lvls, ordered = TRUE)
  dat[,direction:=ifelse(estimate>0,"+","-")]
  
  ggplot(data = dat[modifying_celltype==TRUE,], 
         aes(x=directional_diff,
             y = tail(lvls,1))) +
    #### Lvl 1 ####
    ggridges::stat_density_ridges(
      aes(x = value, 
          y = lvls[1],
          color = lvls[1],
          fill = 0.5 - abs(0.5 - stat(ecdf))) ,  
      data=d_melt, 
      # color="purple",
       geom = "density_ridges_gradient", calc_ecdf = TRUE,
      quantiles = 4, quantile_lines = TRUE, 
      quantile_fun=function(x,...)mean(x,na.rm=TRUE),
      alpha = 0.5) +
   #### Lvl 2 ####
   ggridges::stat_density_ridges(
     aes(x = get(value_var), 
         y = lvls[2],
         color=lvls[2],
         fill = 0.5 - abs(0.5 - stat(ecdf))),
     data = dat, 
      # color="red",
       geom = "density_ridges_gradient", calc_ecdf = TRUE,
      quantiles = 4, quantile_lines = TRUE, 
     quantile_fun=function(x,...)mean(x,na.rm=TRUE),
      alpha = 0.5) + 
  #### Lvl 3 ####
  ggridges::stat_density_ridges(
     aes(x = get(value_var), 
         y = lvls[3],
         color=lvls[3],
         fill = 0.5 - abs(0.5 - stat(ecdf))),
     data = dat[modifying_celltype==FALSE,], 
      # color="red",
       geom = "density_ridges_gradient", calc_ecdf = TRUE,
      quantiles = 4, quantile_lines = TRUE, 
     quantile_fun=function(x,...)mean(x,na.rm=TRUE),
      alpha = 0.5) + 
   #### Lvl 4 ####
    ggridges::stat_density_ridges(
     aes(x = get(value_var),
         y = lvls[4],
         color = lvls[4],
         fill = 0.5 - abs(0.5 - stat(ecdf))),
     data = dat[modifying_celltype==TRUE & p.value<0.05,],  
       geom = "density_ridges_gradient", calc_ecdf = TRUE,
      quantiles = 4, quantile_lines = TRUE, 
      quantile_fun=function(x,...)mean(x,na.rm=TRUE),
      jittered_points = FALSE, 
      position = position_points_jitter(width = 0.05, 
                                        yoffset = -.1,
                                        height = 0),
    point_shape = '|', point_size = 3,alpha = 0.5) +
   #### Lvl 5 ####
    ggridges::stat_density_ridges(
     aes(x = get(value_var), 
         y = lvls[5],
         color=lvls[5],
         fill = 0.5 - abs(0.5 - stat(ecdf))),
      data = dat[modifying_celltype==TRUE & p.value<0.005,],  
      # color="red",
       geom = "density_ridges_gradient", calc_ecdf = TRUE,
      quantiles = 4, quantile_lines = TRUE, 
     quantile_fun=function(x,...)mean(x,na.rm=TRUE),
      alpha = 0.5) + 
    #### Lvl 6 ####
    ggridges::stat_density_ridges(
     aes(x = get(value_var), 
         y = lvls[6],
         color=lvls[6],
         fill = 0.5 - abs(0.5 - stat(ecdf))),
      data = dat[p.value<0.0001,],  
      # color="red",
       geom = "density_ridges_gradient", calc_ecdf = TRUE,
      quantiles = 4, quantile_lines = TRUE, 
     quantile_fun=function(x,...)mean(x,na.rm=TRUE),
      alpha = 0.5) +  
    #### extra ####
   scale_fill_viridis_c(option = "mako", alpha=.8) +
   scale_color_viridis_d(option = "plasma", alpha=.8, end=.8) +
   facet_wrap(facets = variable~.,
               scales="free",
               ncol = 3) +
    labs(y="Density", 
         x=xlabel) + 
    theme_bw() + 
    theme(strip.background = element_blank()) +
    xlim(limits = c(0, NA))
   # geom_boxplot(aes(color=direction),
   #              data = dat[modifying_celltype==TRUE,],
   #               position = position_identity(), 
   #               show.legend = TRUE) + 
   #  stat_boxplot(geom = "vline", aes(xintercept = ..xlower..),
   #               width=position_dodge2(width = 1),
   #               show.legend = FALSE) +
   #  stat_boxplot(geom = "vline", aes(xintercept = ..xmiddle..),
   #               width=position_dodge2(width = 1),
   #               show.legend = FALSE) +
   #  stat_boxplot(geom = "vline", aes(xintercept = ..xupper..),
   #               width=position_dodge2(width = 1),
   #               show.legend = FALSE) 
}

```

```{r}
dplt1 <- plot_distributions(dat = dat,
                            d_melt = d_melt,
                            value_var = "directional_diff")
dplt1
```

```{r}
dplt2 <- plot_distributions(dat = dat,
                            d_melt = d_melt,
                            value_var = "value")
dplt2
```
 

```{r}
model_dt <- res_dt[term=="(Intercept)"]
plot(abs(sig_dt$estimate),
     sig_dt$mdata_diseases)
plot(abs(sig_dt$estimate),
     abs(sig_dt$mdata_mean_value-sig_dt$mdata_mean_value_nonmodifier_celltype))


ggplot(dat,aes(x=CellType,
               y=value,
               fill=CellType)) +
  geom_violin(show.legend = FALSE) +
  facet_wrap(facets = .~variable, 
             scales = "free")
```




# Session info 

<details>
```{r}
sessionInfo()
```
</details>
<hr>