---
title: "neurodegeneration"
author: "Brian M. Schilder"
date: "2023-03-21"
output: html_document
---

Trying to have another look through our results for anything neurodegeneration-related, either at the level of phenotype or disease.


# Import results 

```{r}
res <- MultiEWCE::load_example_results()
annot <- HPOExplorer::load_phenotype_to_genes(3)
```

## Alzheimer's

AD is not enriched for any phenotype, likely due to the small number of genes included. GWAS results are more comprehensive than this.
https://hpo.jax.org/app/browse/term/HP:0002511

```{r}
ad <- res[grepl("alzheimer",Phenotype,ignore.case = TRUE)]
nrow(ad) ## 77 results (1 per celltype)
ad_sig <- ad[q<0.05]
nrow(ad_sig) ## 0 sig results

MultiEWCE::create_dt(ad_sig)
```

## Parkinson's

## Phenotypes

"Parkinson's disease" is not in the HPO, but the branch "Parkinsonism" is.
https://hpo.jax.org/app/browse/term/HP:0001300

We actually do have 2 sig phenotypes here.

```{r}
pd_phenos <- HPOExplorer::make_phenos_dataframe(ancestor="Parkinsonism")
pd1 <- res[HPO_ID %in% pd_phenos$HPO_ID & q<0.05]

MultiEWCE::create_dt(pd1)
```

## Diseases

We can also search for phenotypes via "Parkinson's" as a disease.

This gives us lots of phenotypes, but many of them are not specific to PD: e.g.
*Autosomal recessive inheritance*, 
*Open mouth*, 
*Macrocephaly*,	
*Urinary urgency*

```{r}
d <- annot[grepl("parkinson",DiseaseName, ignore.case = TRUE)]
pd <- res[HPO_ID %in% d$HPO_ID & q<0.05]

MultiEWCE::create_dt(pd)
```

Let's try to narrow it down to very PD-specific phenotypes.

These PD-associated phenotypes only appear once in the HPO annotations. 
Thus, they must be PD-specific.

That leaves us with just 2 phenotypes.

```{r}
n_pd_terms <- length(unique(pd_phenos$HPO_ID))
pd_pheno_freqs <- sort(table(annot[HPO_ID %in% d$HPO_ID]$HPO_ID))
hist(pd_pheno_freqs, 50)
pd_specific <- pd[HPO_ID %in% names(pd_pheno_freqs)[pd_pheno_freqs<=1+n_pd_terms]]

MultiEWCE::create_dt(pd_specific)
```


We can lift this filter a bit and look for PD-associated phenotypes that only appear in 5 or less non-PD diseases. We see phenotype that can indeed be associated with PD, but are not necessarilly exclusive to it: e.g. 
*Diminished movement*, 
*Abnormal aggressive, impulsive or violent behavior*, 
*Abnormal nervous system physiology*

```{r} 
pd_specific2 <- pd[HPO_ID %in% names(pd_pheno_freqs)[pd_pheno_freqs<=5+n_pd_terms]]
MultiEWCE::create_dt(pd_specific2)
```


## Mental deterioration

Get all the phenos from the Mental deterioration branch. This includes many forms of dementia 
 https://hpo.jax.org/app/browse/term/HP:0001268

As we've seen before, only the higher-level term "Mental deterioration" remains significant. Perhaps due to the greater number of genes in this higher level.

```{r}
phenos <- HPOExplorer::make_phenos_dataframe(ancestor="Mental deterioration")
md <- res[HPO_ID %in% phenos$HPO_ID & q<0.05]

MultiEWCE::create_dt(md)
```

## Atrophy/Degeneration affecting the central nervous system

HPO has the term: **Atrophy/Degeneration affecting the central nervous system** 
https://hpo.jax.org/app/browse/term/HP:0007367

In our results, we found a number of significantly enriched phenotypes
that fall under that branch:

```{r}
phenos <- HPOExplorer::make_phenos_dataframe(ancestor="Atrophy/Degeneration affecting the central nervous system")

at <- res[HPO_ID %in% phenos$HPO_ID & q<0.05]
MultiEWCE::create_dt(at)
```

# Multi-trait intersection

## dementia/neurodegeneration vs. diabetes/obesity

Next, let's find examples of celltype-specific enrichment results that relate to diseases/phenotypes in the categories of dementia/neurodegeneration vs. diabetes/obesity.

```{r}

```


# Session info 

<details>
```{r}
sessionInfo()
```
</details>
